---
title: "Integrated serosurveillance to assess disease elimination in coastal Ecuador"
subtitle: "Trachoma antigen longitudinal analysis"
author: "Ben Arnold ben.arnold@ucsf.edu"
date: "updated: `r Sys.time()`"
format:
  html:
    
    toc: true
    embed-resources: true
    code-fold: true
    link-external-newwindow: true
---

# Summary

This script summarizes IgG antibody responses to _Chlaymdia trachomatis_ Pgp3 and Ct684 antigens among children in the ECoMiD birth cohort. 

We then estimate the seroconversion rate (SCR) to Pgp3 for comparison with other global populations along the path toward elimination.  We estimate the SCR using two approaches: a catalytic model based on age-seroprevalence, and non-parametrically using longitudinal seroconversion among children at risk.

# Preamble

```{r preamble}
library(here)
here()
#----------------------------
# source the config file
#----------------------------
source(here("R","0-config.R"))

# unload the raster package to avoid conflicts with dplyr::select
detach("package:raster", unload = TRUE)

```

# Load the clean MBA data

```{r load antibody data}
#-------------------------------
# load the formatted MBA data
# for EcoMID, created by
# 0-all_sero_results_dataclean.Rmd
#
# data are in long format
#-------------------------------
d <- read_rds(here("../../data/serology/final/ecomid_serology_datasets","ecomid_sero_antigen_seropos_long_2025-12-18.rds"))

# drop monitoring targets
d1 <- d %>%
  filter( !antigen %in% c("gst","monitor","vero"))
```

```{r load child visit data}
#-------------------------------
# load child visit data for
# exact dates at age of DBS collection
#-------------------------------
dvisit <- read_rds(file=here("../../data/ecomid-field/final","ecomid_child_visit_info_2025-09-10.rds"))

#-------------------------------
# join visit ages to serology
# data and create a final age
# variable
#-------------------------------
d2 <- d1 %>%
  left_join(dvisit, join_by("community","community_name","childid","timepoint"), relationship = "many-to-one") %>%
  mutate(age_days = case_when(
    !is.na(agedays_dbs) ~ agedays_dbs,
    is.na(agedays_dbs) & !is.na(agedays_stool) ~ agedays_stool,
    is.na(agedays_dbs) & is.na(agedays_stool) ~ agedays_svy,
    TRUE ~ NA
  ),
    # there are 2 children with visits that don't appear in the childvisit data
    # for them, it is both at 12 months, so assume age is DOB + 365 days
    agedays = ifelse(is.na(age_days), difftime(time1=dob, time2=dob + 365, units="days"), age_days), 
  age_months = age_days/30.4,
  age_years = age_days/365.25
           
           )
```

# MFI by age

Individual longitudinal trajectories via spaghetti plot, just to get a feel for the longitudinal patterns


```{r mfi by age pgp3}

# Longitudinal trajectories pgp3
# identify children who were ever Pgp3 positive
d2_pgp3pos <- d2 %>%
  group_by(childid, antigen) %>%
  mutate(everpos = max(seropos, na.rm=TRUE)) %>%
  filter(antigen == "pgp3" & everpos ==1)

# identify when they first became positive to Pgp3
first_pgp3 <- d2_pgp3pos %>%
  filter(seropos==1) %>%
  group_by(childid) %>%
  mutate(firstpos = min(timepoint),
         firstpos = paste0(firstpos," month visit"),
         firstpos = factor(firstpos, levels=paste0(c(6,9,12,18,24)," month visit"))
  ) %>%
  select(childid,firstpos) %>%
  distinct()

nrow(first_pgp3) # n=55 (!)

# join back to data frame
d2_pgp3pos2 <- d2_pgp3pos %>%
  left_join(first_pgp3, join_by(childid))  %>%
  # filter out a few (n=4) observations with missing MFI
  filter(!is.na(mfi))

# grab cutoffs for plotting
pgp3cut  <- log10(d2$mficut[d2$antigen=="pgp3"][1])

# custom log labels
log10labs <- c( 
  expression(10^0),
  expression(10^1),
  expression(10^2),
  expression(10^3),
  expression(10^4),
  expression(10^5)
)

# make plot
plot_mfi_long_pgp3_pos <- ggplot(data=d2_pgp3pos2, aes(x = age_months,y=log10mfi, group = childid)) +
  facet_grid(community_type~firstpos) + 
  geom_line(color=cbpal[7],alpha=0.8) +
  geom_hline(yintercept = pgp3cut, lty="dashed") +
  scale_x_continuous(breaks = c(6,9,12,15,18,21,24)) +
  scale_y_continuous(breaks = 0:5, ) +
  coord_cartesian(ylim=c(0,5)) + 
  labs(x = "Age (months)", y = "log10 Pgp3 IgG Luminex response (MFI-bg)",subtitle="Visit of first seropositivity") +
  # theme_minimal() +
  theme(
    panel.grid.minor.x = element_blank(),
    axis.text.x = element_text(size=8),
    plot.subtitle = element_text(hjust = 0.5),
    strip.text.y = element_text(angle=0)
  )

plot_mfi_long_pgp3_pos



ggsave(filename = here("output","ecomid_mfi_pgp3_by_age_comm.png"),plot_mfi_long_pgp3_pos, device="png",width = 200, height = 180, units = "mm")
```

## Individual child trajectories

Plot Pgp3 and Ct694 for the 30 children who were seropositive to both at some point during follow-up. 

Antibody responses to the two antigens are broadly consistent and seem to track longitudinally with high consistency in most cases. 

```{r mfi by age pgp3 and ct694}
#-------------------------------
# in light of the large number
# of pgp3 positive children above
# examine ct694 as a confirmatory test
#-------------------------------
dct <- d2 %>%
  group_by(childid, antigen) %>%
  mutate(everpos = max(seropos)) %>%
  filter(antigen %in% c("pgp3","ct694"))

#-------------------------------
# pivot the MFI and seropos
# values to wide format by sample
#-------------------------------
dct2 <- dct %>%
  select(community_type,sampleid,childid,timepoint, age_months, antigen,log10mfi,seropos) %>%
  pivot_wider(id_cols=c("community_type","sampleid","childid","timepoint","age_months"),names_from = "antigen",values_from=c("log10mfi","seropos")) %>%
  # create a joint seropositivity indicator
  mutate(seropos_pgp3ct694 = ifelse(seropos_pgp3==1 & seropos_ct694==1,1,0))

# of 104 samples seropositive to Pgp3, 63 were also seropositive to Ct694
# of 65 samples seropositive to Ct694, 63 were also seropositive to Pgp3
table(dct2$seropos_pgp3,dct2$seropos_ct694)


#-------------------------------
# identify when they first became positive to both
# there are 31 children in this
# set
#-------------------------------
firstpos_pgp3ct694 <- dct2 %>%
  filter(seropos_pgp3ct694==1) %>%
  group_by(childid) %>%
  mutate(firstpos = min(timepoint),
         firstpos = paste0("Age ",firstpos," months"),
         firstpos = factor(firstpos, levels=paste0("Age ",c(6,9,12,18,24)," months"))
  ) %>%
  select(childid,firstpos) %>%
  distinct()

# join back to data frame
dct3 <- dct2 %>%
  left_join(first_pgp3, join_by(childid)) %>%
  # create an arbitrary childid variable for plotting (to avoid plotting real IDs)
  group_by(childid) %>%
  mutate(childid_plot = paste0(substr(as.character(community_type),1,1),"-",sprintf("%03s",cur_group_id()))) %>%
  # remove rows that have missing MFI values for the plot
  filter(!is.na(log10mfi_ct694) | !is.na(log10mfi_pgp3))

```



```{r plot ct antigens for children positive to both, fig.height=10}
#-------------------------------
# plot MFI values for the
# 31 children who were seropositive to both
# antigens, Pgp3 and Ct694
#-------------------------------

# grab cutoffs for plotting
pgp3cut  <- log10(d2$mficut[d2$antigen=="pgp3"][1])
ct694cut <- log10(d2$mficut[d2$antigen=="ct694"][1])


plot_pgp3ct694_mfi <- ggplot(dct3 %>% filter(childid %in% firstpos_pgp3ct694$childid), 
                             aes(x=age_months,)) +
  facet_wrap(.~childid_plot) +
  # add cutoffs
  geom_hline(yintercept = pgp3cut, lty="dashed", lwd=0.2, color=cbpal[2]) +
  geom_hline(yintercept = ct694cut, lty="dashed", lwd=0.2, color=cbpal[6]) +

  geom_line(aes(y=log10mfi_pgp3), alpha = 0.5, color=cbpal[7]) +
  geom_line(aes(y=log10mfi_ct694), alpha = 0.5, color=cbpal[6]) +
  scale_x_continuous(breaks = seq(6,24,by=6)) +
  coord_cartesian(xlim=c(6,24),ylim=c(0,4.5)) +
  # scale_color_manual(values=cbpal[2:3]) +
  labs(x = "Age in months", y="log10 IgG Luminex response (MFI-bg)",
       # title = "IgG trajectories for children seropositive to 2 antigens",
       subtitle="Pgp3 (orange) and Ct694 (blue)") +
  theme_minimal()

plot_pgp3ct694_mfi

ggsave(filename = here("output","ecomid-longtrajectories-pgp3ct694pos.png"),plot=plot_pgp3ct694_mfi, device = "png", width=120, height=200, units="mm")

```


Plot Pgp3 and Ct694 for the 24 children who were seropositive to Pgp3 only (not Ct694). 

These longitudinal patterns are also consistent between antibody responses, and although Ct694 does not cross the seropositivity threshold it does track Pgp3 longitudinally in nearly all cases.  

```{r plot ct antigens for children positive to just pgp3, fig.height=8}
#-------------------------------
# plot MFI values for the
# 24 children who were seropositive to only
# Pgp3 (and not Ct694)
#-------------------------------
# identify children just posiitive to Pgp3
justpgp3 <- first_pgp3$childid[!first_pgp3$childid %in% firstpos_pgp3ct694$childid]
length(justpgp3)

# grab cutoffs for plotting
pgp3cut  <- log10(d2$mficut[d2$antigen=="pgp3"][1])
ct694cut <- log10(d2$mficut[d2$antigen=="ct694"][1])

# make plot
plot_pgp3_mfi <- ggplot(dct3 %>% filter(childid %in% justpgp3), 
                             aes(x=age_months,)) +
  facet_wrap(.~childid_plot, ncol=6) +
  # add cutoffs
  geom_hline(yintercept = pgp3cut, lty="dashed", lwd=0.2, color=cbpal[2]) +
  geom_hline(yintercept = ct694cut, lty="dashed", lwd=0.2, color=cbpal[6]) +

  geom_line(aes(y=log10mfi_pgp3), alpha = 0.5, color=cbpal[7]) +
  geom_line(aes(y=log10mfi_ct694), alpha = 0.5, color=cbpal[6]) +
  scale_x_continuous(breaks = seq(6,24,by=6)) +
  coord_cartesian(xlim=c(6,24),ylim=c(0,4.5)) +
  # scale_color_manual(values=cbpal[2:3]) +
  labs(x = "Age in months", y="log10 IgG Luminex response (MFI-bg)",
       # title = "IgG trajectories for children seropositive to just Pgp3",
       subtitle="Pgp3 (orange) and Ct694 (blue)") +
  theme_minimal()

plot_pgp3_mfi

ggsave(filename = here("output","ecomid-longtrajectories-pgp3pos.png"),plot=plot_pgp3_mfi, device = "png", width=120, height=160, units="mm")

```


# Seroprevalence by age


```{r load Pgp3 antibody data}

# filter to Pgp3 responses
d3 <- d2 %>%
  # filter( !antigen %in% c("gst","monitor","vero"))
  filter(antigen=="pgp3") 

```


Average seroprevalence by age

```{r seroprev by age}

#-------------------------------
# estimate seroprevalence 
# by age and community type
#-------------------------------
dprev_age <- d3 %>%
  mutate(community_type2 = case_when(
    community_type %in% c("Rural Road","Rural River") ~ "Rural",
    TRUE ~ as.character(community_type)
    ),
    community_type2 = factor(community_type2,levels=c("Esmeraldas","Borbon","Rural"))
    )%>%
  group_by(community_type2,timepoint) %>%
  summarize(
    nobs = sum(!is.na(seropos)),
    npos = sum(seropos, na.rm=TRUE)
  ) %>%
  rowwise() %>%
  mutate(
    seroprev = (npos/nobs)*100,
    seroprev_min95 = binom.test(x=npos,n=nobs)$conf.int[1]*100,
    seroprev_max95 = binom.test(x=npos,n=nobs)$conf.int[2]*100
  )

kbl(dprev_age, digits=1)

pcols <- RColorBrewer::brewer.pal(8,"Dark2")[3:1]
plot_pgp3_by_age <- ggplot(data = dprev_age, 
                           aes(x = timepoint, 
                               y=seroprev,
                               color=community_type2,
                               fill=community_type2)
                           ) +
  facet_grid(.~community_type2) +
  geom_point() +
  geom_ribbon(aes(ymin=seroprev_min95,ymax=seroprev_max95),alpha=0.2,color=NA) +
  geom_line() +
  scale_x_continuous(breaks = c(0,6,9,12,18,24)) +
  scale_y_continuous(breaks = seq(0,20,by=5)) +
  scale_color_manual(values = pcols) +
  scale_fill_manual(values = pcols) +
  coord_cartesian(xlim=c(6,24)) +
  labs(x = "Age (months)", y = "Seroprevalence (%)", title = "Pgp3 seroprevalence by age, ECoMiD") +
  theme_minimal() +
  theme(
    legend.position = "none",
    panel.grid.minor.x = element_blank()
  )

plot_pgp3_by_age
```



# Estimate SCR longitudinal

Estimate the seroconversion rate using incident seroconversions among children at risk.  Assume seroconversions occur at the mid-point between longitudinal measurements (for calculating person-time at risk). This is unbiased on average and a reasonable/standard approach.  

Bootstrap resample children with replacement to estimate 95% CIs in the incidence rate. 

For very rare events, where the bootstrap distribution is clearly unsmooth, use estimate an exact 95% CI using the Garwood (exact Poisson) confidence interval. 


```{r identify incident seroconversions}

#-------------------------------
# create indicators for whether
# a child is at risk of 
# seroconversion and whether
# they have an incident seroconversion
# and tally person time at risk
#
# then, do the opposite, identifying
# incident seroreversions among
# those who were at risk (seropos)
#-------------------------------
d4 <- d3 %>%
  # group by child and sort by visit
  # to identify incident seroconversions
  # estimate time at risk as diff
  # in age between measurements when at risk
  group_by(childid) %>%
  arrange(childid,timepoint) %>%
  mutate(
    # incident seroconversions
    atrisk = ifelse(seropos == 0, 1, 0),
    lead_seropos = lead(seropos),
    seroconv = ifelse(atrisk == 1 & seropos == 0 & lead_seropos == 1, 1, 0),
    lead_age_years = lead(age_years),
    pt = lead_age_years - age_years,
    # assume seroconversion happened at the mid-point of the interval
    # in cases of seroconversion, reduce person time by 1/2
    pt = ifelse(seroconv==1, pt*0.5,pt),
    # set persontime at risk to zero if the child is not at risk
    pt = ifelse(atrisk == 0, 0, pt),
    
    # incident seroreversions and person-time at risk of reversion
    atriskr = ifelse(seropos == 1, 1, 0),
    serorev = ifelse(atriskr == 1 & seropos == 1 & lead_seropos == 0, 1, 0),
    ptr = lead_age_years - age_years,
    ptr = ifelse(serorev == 1, ptr*0.5, ptr),
    ptr = ifelse(atriskr == 0, 0, ptr)

    )

```

```{r estimate incidence rate}
#-------------------------------
# estimate the incidence rate
# bootstrap resample children
# to estimate the 95% CI
#-------------------------------
est_ir <- function(new_episode,pt) {
  new_episodes <- sum(new_episode, na.rm=TRUE)
  ptar <- sum(pt, na.rm=TRUE)
  ir <- new_episodes/ptar
  return(ir)
}

# observed estimated incidence rate
ir_estimate <- est_ir(new_episode = d4$seroconv, pt = d4$pt)

# boostrap resample
idlist <- unique(d4$childid)
nboot <- 1000
set.seed(19373)
boot_samp <- matrix(sample(idlist, size=length(idlist)*nboot, replace=TRUE ), nrow=length(idlist),ncol=nboot)
ir_boots <- foreach(i = 1:nboot, .combine = rbind) %do% {
  di <- left_join(data.frame(childid = boot_samp[,i]), d4, join_by(childid), relationship = "many-to-many")
  iri <- est_ir(new_episode = di$seroconv, pt = di$pt)
  return(iri)
}
# examine the bootstrap distribution to ensure it is smooth
hist(ir_boots, breaks=40)
# looks good (as expected)
# estimate the percentile 95% CI from the boostrap distribution
ir_min95 <- quantile(ir_boots, prob=0.025)
ir_max95 <- quantile(ir_boots, prob=0.975)

# return results
cat("The longitudinal estimate of the SCR (bootstrap 95% CI) is:\n",
    sprintf("%1.1f",ir_estimate*100)," (",sprintf("%1.1f",ir_min95*100),"to",sprintf("%1.1f",ir_max95*100),") per 100 person-years\n")

# create a simple table of results
scr_tab <- data.frame(
  desc = "Pgp3 seroconversion",
  nchild = length(unique(d4$childid)),
  nseroc = sum(d4$seroconv,na.rm=TRUE),
  ptar = sum(d4$pt,na.rm=TRUE),
  ir = ir_estimate*100,
  ir_min95 = ir_min95*100,
  ir_max95 = ir_max95*100,
  ir_print = paste0(sprintf("%1.1f",ir_estimate*100)," (",sprintf("%1.1f",ir_min95*100)," to ",sprintf("%1.1f",ir_max95*100),")")
)

kbl(scr_tab %>% select(-ir,-ir_min95,-ir_max95), caption = "Summary of incident seroconversions to Pgp3", col.names = c("","N children", "N seroconversions","Years at risk", "SCR (95% CI) per 100 person-years"))
```

## Stratified estimates by community type

```{r stratified incidence estimates}
#-------------------------------
# identify community type from
# sampleid
#-------------------------------
d5 <- d4 %>%
  mutate(community_code = substr(sampleid,1,1),
         community_type = case_when(
           community_code == "E" ~ "Esmeraldas",
           community_code == "B" ~ "Borbon",
           TRUE ~ "Rural villages"),
         community_type = factor(community_type, levels=c("Esmeraldas","Borbon","Rural villages"))
  )

# stratified datasets for convenience
d5_e <- d5 %>% filter(community_type == "Esmeraldas")
d5_b <- d5 %>% filter(community_type == "Borbon")
d5_r <- d5 %>% filter(community_type == "Rural villages")
#-------------------------------
# observed estimated incidence rates
#-------------------------------
ir_estimate_e <- est_ir(new_episode = d5_e$seroconv, pt = d5_e$pt)
ir_estimate_b <- est_ir(new_episode = d5_b$seroconv, pt = d5_b$pt)
ir_estimate_r <- est_ir(new_episode = d5_r$seroconv, pt = d5_r$pt)
```

```{r stratified bootstraps}
#-------------------------------
# bootstrap esmeraldas, borbon and rural
# villages
#-------------------------------


#-------------------------------
# boostrap resample, Borbon
#-------------------------------
idlist <- unique(d5_e$childid)
nboot <- 1000
set.seed(234236)
boot_samp_e <- matrix(sample(idlist, size=length(idlist)*nboot, replace=TRUE ), nrow=length(idlist),ncol=nboot)
ir_boots <- foreach(i = 1:nboot, .combine = rbind) %do% {
  di <- left_join(data.frame(childid = boot_samp[,i]), d5_e, join_by(childid), relationship = "many-to-many")
  iri <- est_ir(new_episode = di$seroconv, pt = di$pt)
  return(iri)
}
# examine the bootstrap distribution to ensure it is smooth
hist(ir_boots, breaks=40)
# looks good (as expected)
# estimate the percentile 95% CI from the boostrap distribution
ir_min95_e <- quantile(ir_boots, prob=0.025)
ir_max95_e <- quantile(ir_boots, prob=0.975)


#-------------------------------
# boostrap resample, Borbon
#-------------------------------
idlist <- unique(d5_b$childid)
nboot <- 1000
set.seed(4578792)
boot_samp_b <- matrix(sample(idlist, size=length(idlist)*nboot, replace=TRUE ), nrow=length(idlist),ncol=nboot)
ir_boots <- foreach(i = 1:nboot, .combine = rbind) %do% {
  di <- left_join(data.frame(childid = boot_samp[,i]), d5_b, join_by(childid), relationship = "many-to-many")
  iri <- est_ir(new_episode = di$seroconv, pt = di$pt)
  return(iri)
}
# examine the bootstrap distribution to ensure it is smooth
hist(ir_boots, breaks=40)
# it is not smooth. So instead estimate the
# exact Garwood 95% CI by inverting the poisson distribution
# estimate the percentile 95% CI from the boostrap distribution
events_b <- sum(d5_b$seroconv, na.rm=TRUE)
pt_b <- sum(d5_b$pt, na.rm=TRUE)
ci_b <- poisson.test(events_b, pt_b, conf.level = 0.95)$conf.int
ir_min95_b <- ci_b[1]
ir_max95_b <- ci_b[2]

#-------------------------------
# boostrap resample, Rural villages
#-------------------------------
idlist <- unique(d5_r$childid)
nboot <- 1000
set.seed(234234)
boot_samp_b <- matrix(sample(idlist, size=length(idlist)*nboot, replace=TRUE ), nrow=length(idlist),ncol=nboot)
ir_boots <- foreach(i = 1:nboot, .combine = rbind) %do% {
  di <- left_join(data.frame(childid = boot_samp[,i]), d5_r, join_by(childid), relationship = "many-to-many")
  iri <- est_ir(new_episode = di$seroconv, pt = di$pt)
  return(iri)
}
# examine the bootstrap distribution to ensure it is smooth
hist(ir_boots, breaks=40)
# looks good (as expected)
# estimate the percentile 95% CI from the boostrap distribution
ir_min95_r <- quantile(ir_boots, prob=0.025)
ir_max95_r <- quantile(ir_boots, prob=0.975)

```

```{r summarize stratified results}
#-------------------------------
# bring all of the information
# together into a single table
#-------------------------------
scr_tab_e <- data.frame(
  desc = "Esmeraldas",
  nchild = length(unique(d5_e$childid)),
  nseroc = sum(d5_e$seroconv,na.rm=TRUE),
  ptar = sum(d5_e$pt,na.rm=TRUE),
  ir = ir_estimate_e*100,
  ir_min95 = ir_min95_e*100,
  ir_max95 = ir_max95_e*100,
  ir_print = paste0(sprintf("%1.1f",ir_estimate_e*100)," (",sprintf("%1.1f",ir_min95_e*100)," to ",sprintf("%1.1f",ir_max95_e*100),")")
)

scr_tab_b <- data.frame(
  desc = "Borbon",
  nchild = length(unique(d5_b$childid)),
  nseroc = sum(d5_b$seroconv,na.rm=TRUE),
  ptar = sum(d5_b$pt,na.rm=TRUE),
  ir = ir_estimate_b*100,
  ir_min95 = ir_min95_b*100,
  ir_max95 = ir_max95_b*100,
  ir_print = paste0(sprintf("%1.1f",ir_estimate_b*100)," (",sprintf("%1.1f",ir_min95_b*100)," to ",sprintf("%1.1f",ir_max95_b*100),")")
)

scr_tab_r <- data.frame(
  desc = "Rural villages",
  nchild = length(unique(d5_r$childid)),
  nseroc = sum(d5_r$seroconv,na.rm=TRUE),
  ptar = sum(d5_r$pt,na.rm=TRUE),
  ir = ir_estimate_r*100,
  ir_min95 = ir_min95_r*100,
  ir_max95 = ir_max95_r*100,
  ir_print = paste0(sprintf("%1.1f",ir_estimate_r*100)," (",sprintf("%1.1f",ir_min95_r*100)," to ",sprintf("%1.1f",ir_max95_r*100),")")
)

scr_tab_all <- bind_rows(scr_tab_e,scr_tab_b,scr_tab_r)

kbl(scr_tab_all %>% select(-ir,-ir_min95,-ir_max95), caption = "Summary of incident seroconversions to Pgp3", col.names = c("","N children", "N seroconversions","Years at risk", "SCR (95% CI) per 100 person-years"))

#-------------------------------
# save the results to a csv 
# file for future use
# include Overall as first row!
#-------------------------------
scr_res <- bind_rows(scr_tab, scr_tab_all)
write_csv(scr_res, file = here("output","ecomid_long_scr_by_community_pgp3.csv"))
```

# Estimate SRR longitudinal

## Overall rate, 6-24m
Since there are so few periods at risk of sero-reversion, first estimate the overall rate

```{r}
#| label: estimate longitudinal SRR 6 to 24m

#-------------------------------
# estimate the seroreversion rate
# bootstrap resample children
# to estimate the 95% CI
#-------------------------------

# observed estimated seroreversion rate
srr_estimate <- est_ir(new_episode = d4$serorev, pt = d4$ptr)

# boostrap resample
idlist <- unique(d4$childid)
nboot <- 1000
set.seed(19373)
boot_samp <- matrix(sample(idlist, size=length(idlist)*nboot, replace=TRUE ), nrow=length(idlist),ncol=nboot)
srr_boots <- foreach(i = 1:nboot, .combine = rbind) %do% {
  di <- left_join(data.frame(childid = boot_samp[,i]), d4, join_by(childid), relationship = "many-to-many")
  iri <- est_ir(new_episode = di$serorev, pt = di$ptr)
  return(iri)
}
# examine the bootstrap distribution to ensure it is smooth
hist(srr_boots, breaks=40)
# looks good (as expected)
# estimate the percentile 95% CI from the boostrap distribution
srr_min95 <- quantile(srr_boots, prob=0.025)
srr_max95 <- quantile(srr_boots, prob=0.975)

# return results
cat("The longitudinal estimate of the SRR (bootstrap 95% CI) is:\n",
    sprintf("%1.1f",srr_estimate*100)," (",sprintf("%1.1f",srr_min95*100),"to",sprintf("%1.1f",srr_max95*100),") per 100 person-years\n")

# create a simple table of results
srr_tab <- data.frame(
  desc = "Pgp3 seroreversion 6-24m",
  nchild = length(unique(d4$childid)),
  nseror = sum(d4$serorev,na.rm=TRUE),
  ptarr = sum(d4$ptr,na.rm=TRUE),
  srr = srr_estimate*100,
  srr_min95 = srr_min95*100,
  srr_max95 = srr_max95*100,
  srr_print = paste0(sprintf("%1.1f",srr_estimate*100)," (",sprintf("%1.1f",srr_min95*100)," to ",sprintf("%1.1f",srr_max95*100),")")
)

kbl(srr_tab %>% select(-srr,-srr_min95,-srr_max95), caption = "Summary of incident seroreversions to Pgp3, 6-24m", col.names = c("","N children", "N seroreversions","Years at risk", "SRR (95% CI) per 100 person-years"))



```

## Rate 12-24m

Examination of longitudinal trajectories of children suggests that some sero-reversion could be due to loss of maternal IgG during the first 12 months.  Estimate the seroreversion rate among children who were Pgp3 positive at 12 or 18 months. 


```{r}
#| label: estimate longitudinal SRR 12 to 24m

#-------------------------------
# estimate the seroreversion rate
# bootstrap resample children
# to estimate the 95% CI
#
# limit the data to 12-24m
#-------------------------------

d4_12to24m <- d4 %>%
  filter(timepoint %in% c(12, 18, 24))

# observed estimated seroreversion rate
srr_estimate <- est_ir(new_episode = d4_12to24m$serorev, pt = d4_12to24m$ptr)

# boostrap resample
idlist <- unique(d4_12to24m$childid)
nboot <- 1000
set.seed(19373)
boot_samp <- matrix(sample(idlist, size=length(idlist)*nboot, replace=TRUE ), nrow=length(idlist),ncol=nboot)
srr_boots <- foreach(i = 1:nboot, .combine = rbind) %do% {
  di <- left_join(data.frame(childid = boot_samp[,i]), d4_12to24m, join_by(childid), relationship = "many-to-many")
  iri <- est_ir(new_episode = di$serorev, pt = di$ptr)
  return(iri)
}
# examine the bootstrap distribution to ensure it is smooth
hist(srr_boots, breaks=40)
# looks good (as expected)
# estimate the percentile 95% CI from the boostrap distribution
srr_min95 <- quantile(srr_boots, prob=0.025)
srr_max95 <- quantile(srr_boots, prob=0.975)

# return results
cat("The longitudinal estimate of the SRR 12-24m (bootstrap 95% CI) is:\n",
    sprintf("%1.1f",srr_estimate*100)," (",sprintf("%1.1f",srr_min95*100),"to",sprintf("%1.1f",srr_max95*100),") per 100 person-years\n")

# create a simple table of results
srr_tab_12to24 <- data.frame(
  desc = "Pgp3 seroreversion 12-24m",
  nchild = length(unique(d4_12to24m$childid)),
  nseror = sum(d4_12to24m$serorev,na.rm=TRUE),
  ptarr = sum(d4_12to24m$ptr,na.rm=TRUE),
  srr = srr_estimate*100,
  srr_min95 = srr_min95*100,
  srr_max95 = srr_max95*100,
  srr_print = paste0(sprintf("%1.1f",srr_estimate*100)," (",sprintf("%1.1f",srr_min95*100)," to ",sprintf("%1.1f",srr_max95*100),")")
)

kbl(srr_tab_12to24 %>% select(-srr,-srr_min95,-srr_max95), caption = "Summary of incident seroreversions to Pgp3, 12-24m", col.names = c("","N children", "N seroreversions","Years at risk", "SRR (95% CI) per 100 person-years"))


#-------------------------------
# save the results to a csv 
# file for future use
#-------------------------------
srr_res <- bind_rows(srr_tab, srr_tab_12to24)
write_csv(srr_res, file = here("output","ecomid_long_srr_pgp3.csv"))
```

# Sensitivity analysis: SCR Pgp3+Ct694

As a sensitivity analysis, estimate the seroconversion rate requiring a child to be positive to both Pgp3 and Ct694. Many global monitoring studies rely on Pgp3 alone, but requiring a sample to be positive to both antigens should (in theory) increase the specificity of the assay.

By restricting the definition of seropositivity and seroconversion to samples positive to both Pgp3 and Ct694, we also rule out a couple of examples of "weak" seroconversions where a child's IgG to Pgp3 just barely crosses the cutoff, which could be due to chance. 

```{r }
#| label: identify incident seroconversions pgp3 and ct694

#-------------------------------
# create indicators for whether
# a child is at risk of 
# seroconversion and whether
# they have an incident seroconversion
# and tally person time at risk
#
# then, do the opposite, identifying
# incident seroreversions among
# those who were at risk (seropos)
#-------------------------------

d6 <- d2 %>%
  # filter to Pgp3 and Ct694
  filter(antigen %in% c("pgp3","ct694")) %>%
  select(community_type,childid,timepoint,age_years,sampleid,antigen,seropos) %>%
  pivot_wider(id_cols = c(community_type,childid,timepoint,age_years,sampleid),names_from = "antigen", values_from = "seropos") %>%
  mutate(seropos = ifelse(pgp3==1 & ct694==1,1,0),
         antigen = "pgp3-ct694") %>%
  # group by child and sort by visit
  # to identify incident seroconversions
  # estimate time at risk as diff
  # in age between measurements when at risk
  group_by(childid) %>%
  arrange(childid,timepoint) %>%
  mutate(
    # incident seroconversions
    atrisk = ifelse(seropos == 0, 1, 0),
    lead_seropos = lead(seropos),
    seroconv = ifelse(atrisk == 1 & seropos == 0 & lead_seropos == 1, 1, 0),
    lead_age_years = lead(age_years),
    pt = lead_age_years - age_years,
    # assume seroconversion happened at the mid-point of the interval
    # in cases of seroconversion, reduce person time by 1/2
    pt = ifelse(seroconv==1, pt*0.5,pt),
    # set persontime at risk to zero if the child is not at risk
    pt = ifelse(atrisk == 0, 0, pt)
    )

```

```{r incidence estimates pgp3ct694}

# observed estimated incidence rate
ir_estimate <- est_ir(new_episode = d6$seroconv, pt = d6$pt)

# boostrap resample
idlist <- unique(d6$childid)
nboot <- 1000
set.seed(357295)
boot_samp <- matrix(sample(idlist, size=length(idlist)*nboot, replace=TRUE ), nrow=length(idlist),ncol=nboot)
ir_boots <- foreach(i = 1:nboot, .combine = rbind) %do% {
  di <- left_join(data.frame(childid = boot_samp[,i]), d6, join_by(childid), relationship = "many-to-many")
  iri <- est_ir(new_episode = di$seroconv, pt = di$pt)
  return(iri)
}
# examine the bootstrap distribution to ensure it is smooth
hist(ir_boots, breaks=40)
# looks good (as expected)
# estimate the percentile 95% CI from the boostrap distribution
ir_min95 <- quantile(ir_boots, prob=0.025)
ir_max95 <- quantile(ir_boots, prob=0.975)

# return results
cat("The longitudinal estimate of the SCR using Pgp3+Ct694 (bootstrap 95% CI) is:\n",
    sprintf("%1.1f",ir_estimate*100)," (",sprintf("%1.1f",ir_min95*100),"to",sprintf("%1.1f",ir_max95*100),") per 100 person-years\n")

# create a simple table of results
scr_tab <- data.frame(
  desc = "Pgp3 and Ct694 seroconversion",
  nchild = length(unique(d6$childid)),
  nseroc = sum(d6$seroconv,na.rm=TRUE),
  ptar = sum(d6$pt,na.rm=TRUE),
  ir = ir_estimate*100,
  ir_min95 = ir_min95*100,
  ir_max95 = ir_max95*100,
  ir_print = paste0(sprintf("%1.1f",ir_estimate*100)," (",sprintf("%1.1f",ir_min95*100)," to ",sprintf("%1.1f",ir_max95*100),")")
)

kbl(scr_tab %>% select(-ir,-ir_min95,-ir_max95), caption = "Summary of incident seroconversions to Pgp3+Ct694", col.names = c("","N children", "N seroconversions","Years at risk", "SCR (95% CI) per 100 person-years"))
```

## Stratified estimates by community

```{r stratified incidence estimates pgp3ct694}
#-------------------------------
# identify community type from
# sampleid
#-------------------------------
d7 <- d6 %>%
  mutate(community_code = substr(sampleid,1,1),
         community_type = case_when(
           community_code == "E" ~ "Esmeraldas",
           community_code == "B" ~ "Borbon",
           TRUE ~ "Rural villages"),
         community_type = factor(community_type, levels=c("Esmeraldas","Borbon","Rural villages"))
  )

# stratified datasets for convenience
d7_e <- d7 %>% filter(community_type == "Esmeraldas")
d7_b <- d7 %>% filter(community_type == "Borbon")
d7_r <- d7 %>% filter(community_type == "Rural villages")
#-------------------------------
# observed estimated incidence rates
#-------------------------------
ir_estimate_e <- est_ir(new_episode = d7_e$seroconv, pt = d7_e$pt)
ir_estimate_b <- est_ir(new_episode = d7_b$seroconv, pt = d7_b$pt)
ir_estimate_r <- est_ir(new_episode = d7_r$seroconv, pt = d7_r$pt)
```

```{r stratified bootstraps pgp3ct694 }
#-------------------------------
# bootstrap esmeraldas, borbon and rural
# villages
#-------------------------------

#-------------------------------
# boostrap resample, Esmeraldas
#-------------------------------
idlist <- unique(d7_e$childid)
nboot <- 1000
set.seed(234236)
boot_samp_e <- matrix(sample(idlist, size=length(idlist)*nboot, replace=TRUE ), nrow=length(idlist),ncol=nboot)
ir_boots <- foreach(i = 1:nboot, .combine = rbind) %do% {
  di <- left_join(data.frame(childid = boot_samp[,i]), d7_e, join_by(childid), relationship = "many-to-many")
  iri <- est_ir(new_episode = di$seroconv, pt = di$pt)
  return(iri)
}
# examine the bootstrap distribution to ensure it is smooth
hist(ir_boots, breaks=40)
# it is not smooth. So instead estimate the
# exact Garwood 95% CI by inverting the poisson distribution
# estimate the percentile 95% CI from the boostrap distribution
events_e <- sum(d7_e$seroconv, na.rm=TRUE)
pt_e <- sum(d7_e$pt, na.rm=TRUE)
ci_e <- poisson.test(events_e, pt_e, conf.level = 0.95)$conf.int
ir_min95_e <- ci_e[1]
ir_max95_e <- ci_e[2]


#-------------------------------
# boostrap resample, Borbon
#-------------------------------
idlist <- unique(d7_b$childid)
nboot <- 1000
set.seed(4578792)
boot_samp_b <- matrix(sample(idlist, size=length(idlist)*nboot, replace=TRUE ), nrow=length(idlist),ncol=nboot)
ir_boots <- foreach(i = 1:nboot, .combine = rbind) %do% {
  di <- left_join(data.frame(childid = boot_samp[,i]), d7_b, join_by(childid), relationship = "many-to-many")
  iri <- est_ir(new_episode = di$seroconv, pt = di$pt)
  return(iri)
}
# examine the bootstrap distribution to ensure it is smooth
hist(ir_boots, breaks=40)
# it is not smooth. So instead estimate the
# exact Garwood 95% CI by inverting the poisson distribution
# estimate the percentile 95% CI from the boostrap distribution
events_b <- sum(d7_b$seroconv, na.rm=TRUE)
pt_b <- sum(d7_b$pt, na.rm=TRUE)
ci_b <- poisson.test(events_b, pt_b, conf.level = 0.95)$conf.int
ir_min95_b <- ci_b[1]
ir_max95_b <- ci_b[2]

#-------------------------------
# boostrap resample, Rural villages
#-------------------------------
idlist <- unique(d7_r$childid)
nboot <- 1000
set.seed(234234)
boot_samp_b <- matrix(sample(idlist, size=length(idlist)*nboot, replace=TRUE ), nrow=length(idlist),ncol=nboot)
ir_boots <- foreach(i = 1:nboot, .combine = rbind) %do% {
  di <- left_join(data.frame(childid = boot_samp[,i]), d7_r, join_by(childid), relationship = "many-to-many")
  iri <- est_ir(new_episode = di$seroconv, pt = di$pt)
  return(iri)
}
# examine the bootstrap distribution to ensure it is smooth
hist(ir_boots, breaks=40)
# looks good (as expected)
# estimate the percentile 95% CI from the boostrap distribution
ir_min95_r <- quantile(ir_boots, prob=0.025)
ir_max95_r <- quantile(ir_boots, prob=0.975)

```


```{r summarize stratified results pgp3ct694}
#-------------------------------
# bring all of the information
# together into a single table
#-------------------------------
scr_tab_e <- data.frame(
  desc = "Esmeraldas",
  nchild = length(unique(d7_e$childid)),
  nseroc = sum(d7_e$seroconv,na.rm=TRUE),
  ptar = sum(d7_e$pt,na.rm=TRUE),
  ir = ir_estimate_e*100,
  ir_min95 = ir_min95_e*100,
  ir_max95 = ir_max95_e*100,
  ir_print = paste0(sprintf("%1.1f",ir_estimate_e*100)," (",sprintf("%1.1f",ir_min95_e*100)," to ",sprintf("%1.1f",ir_max95_e*100),")")
)

scr_tab_b <- data.frame(
  desc = "Borbon",
  nchild = length(unique(d7_b$childid)),
  nseroc = sum(d7_b$seroconv,na.rm=TRUE),
  ptar = sum(d7_b$pt,na.rm=TRUE),
  ir = ir_estimate_b*100,
  ir_min95 = ir_min95_b*100,
  ir_max95 = ir_max95_b*100,
  ir_print = paste0(sprintf("%1.1f",ir_estimate_b*100)," (",sprintf("%1.1f",ir_min95_b*100)," to ",sprintf("%1.1f",ir_max95_b*100),")")
)

scr_tab_r <- data.frame(
  desc = "Rural villages",
  nchild = length(unique(d7_r$childid)),
  nseroc = sum(d7_r$seroconv,na.rm=TRUE),
  ptar = sum(d7_r$pt,na.rm=TRUE),
  ir = ir_estimate_r*100,
  ir_min95 = ir_min95_r*100,
  ir_max95 = ir_max95_r*100,
  ir_print = paste0(sprintf("%1.1f",ir_estimate_r*100)," (",sprintf("%1.1f",ir_min95_r*100)," to ",sprintf("%1.1f",ir_max95_r*100),")")
)

scr_tab_pgp3ct694 <- bind_rows(scr_tab_e,scr_tab_b,scr_tab_r)

kbl(scr_tab_pgp3ct694 %>% select(-ir,-ir_min95,-ir_max95), caption = "Summary of incident seroconversions to Pgp3+Ct694", col.names = c("","N children", "N seroconversions","Years at risk", "SCR (95% CI) per 100 person-years"))

#-------------------------------
# save the results to a csv 
# file for future use
# include Overall as first row!
#-------------------------------
scr_res <- bind_rows(scr_tab, scr_tab_pgp3ct694)
write_csv(scr_res, file = here("output","ecomid_long_scr_by_community_pgp3ct694.csv"))
```


# Session Info
```{r session info}
sessionInfo()
```


