---
title: "EcoMID serology"
subtitle: "Pgp3 seroprevalence at 6m and cumulative incidence"
author: "Ben Arnold ben.arnold@ucsf.edu"
date: "updated: `r Sys.time()`"
format:
  html:
    
    toc: true
    embed-resources: true
    code-fold: true
    link-external-newwindow: true
---

# Summary

This script summarizes IgG antibody responses to _Chlaymdia trachomatis_ Pgp3 and Ct684 antigens among children in the ECoMiD birth cohort. 

We then estimate the seroconversion rate (SCR) to Pgp3 for comparison with other global populations along the path toward elimination.  We estimate the SCR using two approaches: a catalytic model based on age-seroprevalence, and non-parametrically using longitudinal seroconversion among children at risk.

# Preamble

```{r preamble}
library(here)
here()
#----------------------------
# source the config file
#----------------------------
source(here("R","0-config.R"))

# unload the raster package to avoid conflicts with dplyr::select
detach("package:raster", unload = TRUE)

```

# Load datasets

## Luminex data

```{r load antibody data}
#-------------------------------
# load the formatted MBA data
# for EcoMID, created by
# 0-all_sero_results_dataclean.Rmd
#
# data are in long format
#-------------------------------
d <- read_rds(here("../../data/serology/final/ecomid_serology_datasets","ecomid_sero_antigen_seropos_long_2025-12-18.rds"))

# drop monitoring targets
d1 <- d %>%
  filter( !antigen %in% c("gst","monitor","vero"))
```

## Survey and household data

```{r load child visit data}
#-------------------------------
# load child visit data for
# exact dates at age of DBS collection
#-------------------------------
dvisit <- read_rds(file=here("../../data/ecomid-field/final","ecomid_child_visit_info_2025-09-10.rds"))

#-------------------------------
# load household characteristics
# to use type of delivery
#-------------------------------
dhh <- read_csv(file = here("../../data/ecomid-field/final","surveydata_20251003.csv")) %>%
  mutate(childid = paste0(community,"B",str_pad(house_ID,4,pad="0"))) %>%
  select(childid,vagbirth)

#-------------------------------
# join visit ages to serology
# data and create a final age
# variable
# also join delivery type
#-------------------------------
d2 <- d1 %>%
  left_join(dhh, join_by("childid")) %>%
  left_join(dvisit, join_by("community","community_name","childid","timepoint"), relationship = "many-to-one") %>%
  mutate(age_days = case_when(
    !is.na(agedays_dbs) ~ agedays_dbs,
    is.na(agedays_dbs) & !is.na(agedays_stool) ~ agedays_stool,
    is.na(agedays_dbs) & is.na(agedays_stool) ~ agedays_svy,
    TRUE ~ NA
  ),
    # there are 2 children with visits that don't appear in the childvisit data
    # for them, it is both at 12 months, so assume age is DOB + 365 days
    agedays = ifelse(is.na(age_days), difftime(time1=dob, time2=dob + 365, units="days"), age_days), 
  age_months = age_days/30.4,
  age_years = age_days/365.25
           
           )
```



# Seroprevalence by age

```{r load Pgp3 antibody data}

# filter to Pgp3 responses
d3 <- d2 %>%
  filter(antigen=="pgp3") 

```


Seroprevalence by age

```{r seroprev by age}
#-------------------------------
# estimate seroprevalence 
# by age and community type
#-------------------------------
dprev_age <- d3 %>%
  mutate(community_type2 = case_when(
    community_type %in% c("Rural Road","Rural River") ~ "Rural",
    TRUE ~ as.character(community_type)
    ),
    community_type2 = factor(community_type2,levels=c("Esmeraldas","Borbon","Rural"))
    )%>%
  group_by(community_type2,timepoint) %>%
  summarize(
    nobs = sum(!is.na(seropos)),
    npos = sum(seropos, na.rm=TRUE)
  ) %>%
  rowwise() %>%
  mutate(
    seroprev = (npos/nobs)*100,
    seroprev_min95 = binom.test(x=npos,n=nobs)$conf.int[1]*100,
    seroprev_max95 = binom.test(x=npos,n=nobs)$conf.int[2]*100
  )

kbl(dprev_age, digits=1)

pcols <- RColorBrewer::brewer.pal(8,"Dark2")[3:1]
plot_pgp3_by_age <- ggplot(data = dprev_age, 
                           aes(x = timepoint, 
                               y=seroprev,
                               color=community_type2,
                               fill=community_type2)
                           ) +
  facet_grid(.~community_type2) +
  geom_point() +
  geom_ribbon(aes(ymin=seroprev_min95,ymax=seroprev_max95),alpha=0.2,color=NA) +
  geom_line() +
  scale_x_continuous(breaks = c(0,6,9,12,18,24)) +
  scale_y_continuous(breaks = seq(0,20,by=5)) +
  scale_color_manual(values = pcols) +
  scale_fill_manual(values = pcols) +
  coord_cartesian(xlim=c(6,24)) +
  labs(x = "Age (months)", y = "Seroprevalence (%)", title = "Pgp3 seroprevalence by age, ECoMiD") +
  theme_minimal() +
  theme(
    legend.position = "none",
    panel.grid.minor.x = element_blank()
  )

plot_pgp3_by_age
```

```{r stratified incidence estimates}
#-------------------------------
# simplified community type
#-------------------------------
d4 <- d3 %>%
  mutate(community_code = substr(sampleid,1,1),
         community_type = case_when(
           community_code == "E" ~ "Esmeraldas",
           community_code == "B" ~ "Borbon",
           TRUE ~ "Rural villages"),
         community_type = factor(community_type, levels=c("Esmeraldas","Borbon","Rural villages"))
  )

#---------------------------------
# Identify children at risk at
# age 6 months
# and indicators of ever seropositive
#---------------------------------
dcum <- d4 %>%
  group_by(childid) %>%
  mutate(pos6m = ifelse(timepoint == 6 & seropos == 1, 1, 0),
         pos6m = max(pos6m, na.rm = TRUE),
         everpos = max(seropos, na.rm = TRUE),
         nobs = n(),
         lastobs = max(timepoint)
) %>%
  select(community_type, vagbirth, childid, pos6m, everpos, nobs, lastobs) %>%
  slice(1)


```

# Ever Pgp3 positive

Estimate the proportion of children who were ever Pgp3 positive, by community type. 
```{r}
#| label: ever positive to Pgp3
#| 
#---------------------------------
# Estimate the proportion of children
# who were ever positive to Pgp3
#
# overall and by community type
#---------------------------------
est_everpos_overall <- dcum %>%
  ungroup() %>%
  summarize(nchild = sum(!is.na(everpos)),
            npos = sum(everpos, na.rm=TRUE)) %>%
  rowwise() %>%
  mutate(pgp3pos = npos/nchild*100,
         pgp3pos_min95 = binom.test(x=npos,n=nchild)$conf.int[1]*100,
         pgp3pos_max95 = binom.test(x=npos,n=nchild)$conf.int[2]*100
        ) %>%
  mutate(table_label = "Overall")

est_everpos_comm <- dcum %>%
  group_by(community_type) %>%
  summarize(nchild = sum(!is.na(everpos)),
            npos = sum(everpos, na.rm=TRUE)) %>%
  rowwise() %>%
  mutate(pgp3pos = npos/nchild*100,
         pgp3pos_min95 = binom.test(x=npos,n=nchild)$conf.int[1]*100,
         pgp3pos_max95 = binom.test(x=npos,n=nchild)$conf.int[2]*100
        ) %>%
  rename(table_label = community_type)

est_everpos <- bind_rows(est_everpos_overall, est_everpos_comm) %>%
  select(table_label, everything())

kbl(est_everpos, digits = 1, caption = "Proportion of children who were ever seropositive to Pgp3 between ages 6 and 24 months overall and by community type") %>%
  kableExtra::group_rows(group_label = "Overall",start_row=1,end_row=1) %>%
  kableExtra::group_rows(group_label = "Community type",start_row=2,end_row=4) %>%
  kableExtra::kable_styling(bootstrap_options = "striped")

```


# Seroprevalence at 6m

Estimate Pgp3 seroprevalence at 6m, stratified by community type

```{r}
#| label: estimate  Pgp3 seroprevalence at 6m

#---------------------------------
# estimate % positive at 6 months
# overall
# by community type
# by birth delivery type
#---------------------------------
est_pos6m_overall <- dcum %>%
  ungroup() %>%
  summarize(nchild = sum(!is.na(pos6m)),
            npos = sum(pos6m, na.rm=TRUE)) %>%
  rowwise() %>%
  mutate(prev6m = npos/nchild*100,
         prev6m_min95 = binom.test(x=npos,n=nchild)$conf.int[1]*100,
         prev6m_max95 = binom.test(x=npos,n=nchild)$conf.int[2]*100
        ) %>%
  mutate(table_label = "Overall")

est_pos6m_comm <- dcum %>%
  group_by(community_type) %>%
  summarize(nchild = sum(!is.na(pos6m)),
            npos = sum(pos6m, na.rm=TRUE)) %>%
  rowwise() %>%
  mutate(prev6m = npos/nchild*100,
         prev6m_min95 = binom.test(x=npos,n=nchild)$conf.int[1]*100,
         prev6m_max95 = binom.test(x=npos,n=nchild)$conf.int[2]*100
        ) %>%
  rename(table_label = community_type)

est_pos6m_birth <- dcum %>%
  group_by(vagbirth) %>%
  summarize(nchild = sum(!is.na(pos6m)),
            npos = sum(pos6m, na.rm=TRUE)) %>%
  rowwise() %>%
  mutate(prev6m = npos/nchild*100,
         prev6m_min95 = binom.test(x=npos,n=nchild)$conf.int[1]*100,
         prev6m_max95 = binom.test(x=npos,n=nchild)$conf.int[2]*100
        ) %>%
  mutate(table_label = ifelse(vagbirth == 1, "Vaginal birth","Cesarean section"))

est_pos6m <- bind_rows(est_pos6m_overall, est_pos6m_birth,est_pos6m_comm) %>%
  select(-vagbirth) %>%
  select(table_label, everything())

kbl(est_pos6m, digits = 1, caption = "Pgp3 seroprevalence at age 6 months overall, by type of birth delivery, and by community type") %>%
  kableExtra::group_rows(group_label = "Overall",start_row=1,end_row=1) %>%
    kableExtra::group_rows(group_label = "Birth type",start_row=2,end_row=3) %>%
  kableExtra::group_rows(group_label = "Community type",start_row=4,end_row=6) %>%
  kableExtra::kable_styling(bootstrap_options = "striped")
```

# Cumulative incidence 6-24m
```{r}
#| label: cumulative incidence estimates
#| 
#---------------------------------
# Among those at risk, determine
# if they had an incident seroconversion
# by age 24 months
#
# overall and by community type
# 
# note, these estimates are approximate
# because only about 80% of children
# completed 24 months of follow-up
#---------------------------------
est_cuminc_overall <- dcum %>%
  filter(pos6m == 0) %>%
  ungroup() %>%
  summarize(nchild = sum(!is.na(everpos)),
            npos = sum(everpos, na.rm=TRUE)) %>%
  rowwise() %>%
  mutate(cuminc = npos/nchild*100,
         cuminc_min95 = binom.test(x=npos,n=nchild)$conf.int[1]*100,
         cuminc_max95 = binom.test(x=npos,n=nchild)$conf.int[2]*100
        ) %>%
  mutate(table_label = "Overall")

est_cuminc_comm <- dcum %>%
  filter(pos6m == 0) %>%
  group_by(community_type) %>%
  summarize(nchild = sum(!is.na(everpos)),
            npos = sum(everpos, na.rm=TRUE)) %>%
  rowwise() %>%
  mutate(cuminc = npos/nchild*100,
         cuminc_min95 = binom.test(x=npos,n=nchild)$conf.int[1]*100,
         cuminc_max95 = binom.test(x=npos,n=nchild)$conf.int[2]*100
        ) %>%
  rename(table_label = community_type)

est_cuminc <- bind_rows(est_cuminc_overall, est_cuminc_comm) %>%
  select(table_label, everything())

kbl(est_cuminc, digits = 1, caption = "Cumulative incidence of seroconversion to Pgp3 between ages 6 and 24 months overall and by community type") %>%
  kableExtra::group_rows(group_label = "Overall",start_row=1,end_row=1) %>%
  kableExtra::group_rows(group_label = "Community type",start_row=2,end_row=4) %>%
  kableExtra::kable_styling(bootstrap_options = "striped")

```

# Session Info
```{r session info}
sessionInfo()
```


